Stability of variational quantum dynamics with Gaussian approximants

### 2013-08-27 paper.tex
writing

\beginsection{Convergence of discretely sampled expansions}

@@

\topinsert\XeTeXpicfile converge.pdf width \hsize
Convergence of a sampled vacuum state, as a function of the number of sample amplitudes.  The solid line shows sampling on a regular grid, the dashed one at random amplitudes.  Details in text.  The dotted line shows the difference between the expanded state and a vacuum state truncated to the extent of the grid; the norm of the truncation error is around $10^{-3}$.
\endinsert

\beginsection{Regularisation of the truncated expansion problem}
%% Picard plots

@@

To capture the concept of a smooth expansion over an irregular set of nonorthogonal vectors, we'll define a quantity called the sampling tension.  Suppose that $|u〉$ and $|v〉$ are nearly parallel vectors, and the state $|ψ〉$ is expanded over a set that includes them, as $|ψ〉≈c_u|u〉+c_v|v〉+⋯$.  A relaxed expansion is one where $c_u|u〉$ and $c_v|v〉$ are as similar as possible; this can be ensured, regardless whether the $c$ are complex numbers of whether $|u〉$ and $|v〉$ are parallel or antiparallel, by requiring that the component of $c_u|u〉$ along $|v〉$ is equal to the component of $c_v|v〉$ along $|u〉$.  The difference between these components is
$$T_{uv}=\left| {c_v〈u|v〉\over ‖|u〉‖²} - {c_u〈v|u〉\over ‖|v〉‖²}\right|.$$
These differences form a tension matrix.  In the case that $|u〉$ and $|v〉$ are nearly orthogonal, their components should be independent.  In this case, the $T_{uv}$ will be small, because of the proportionality to $〈u|v〉$.  The tension matrix is a reasonable measure of the smoothness of an expansion: its Frobenius norm, the sum in quadrature of the components, would be a good summary.

@@

\beginsection{Discrete phase space}

%% Truncated Hilbert space

I still haven't thought in detail about the results with truncated Fock spaces.  Obviously, when we restrict the domain of $G$ to the lowest $N$ energy levels, the restricted operator has at most $N$ singular directions.  Since the low energy levels are the ones a restricted grid tends to represent, those singular directions will have relatively large singular values.  So when we do this, we get a well-posed problem, but we lose the ability to add components with small singular values, in order to expand the states we'll get in the future.  And in fact, by solving the inverse problem, we can always get a nice solution in any case.  Numerical experiments suggest that the variational problem is quite well posed for a single step, too.

Truncating Fock space has a similar effect to truncating the singular values, assuming the singular directions belonging to small values have large particle numbers.  And indeed, the trajectories for the average amplitude change in similar ways due to truncation: they pick up an extra loop in the complex plane.

As the Fock space expands with $N»2R$, there are still only $2R$ singular directions.  Presumably these converge to limits as $N$ increases.  There might be a norm on the matrices $〈N|Dψ〉$, where the truncated elements replaced by zero, that measures this.

>>>>

\beginsection{Discretely sampled expansions}

On a computer, the expansion in phase space must be discretised.  The obvious way to do so is to replace the integral with a sum, so that
$$|ψ〉=d₁|α₁〉+⋯+d_R|α_R〉=|A〉d,$$
where $|A〉=\pmatrix{|α₁〉&⋯&|α_R〉}$, and $d$ is a column vector of the coefficients.
In this paper, $c$ will usually denote the coefficients of expansion over orthogonal bases, and $d$ will represent coefficients of nonorthogonal expansions.  If the system has only a few modes, so that $α$ is a low-dimensional vector, the set of sampled amplitudes $A=\{α₁, …,α_R\}$ could be a regular grid; this makes many things easier.  When there are many modes, and $α$ is a large vector, the storage required for grids becomes exponentially large, and the sample points will have to be irregular.

This expansion is formally equivalent to the continuous one, in that
$$∑_r d_r|α_r〉=|Γ〉f_A,$$
where
$$f_A(β)=π^½∑_rd_re^{½|α_r|²}δ(β-α_r).$$
I don't know enough about generalised functions to be comfortable working with these forms; in particular, the norm of $f_A$ is a bit mysterious.  If this could be clarified, it might provide some useful insights, for example how the norm of $f_A$ compares to that of $〈Γ⁺|ψ〉$.

@@

\topinsert\XeTeXpicfile converge.pdf width \hsize
Convergence of a sampled vacuum state, as a function of the number of sample amplitudes.  The solid line shows sampling on a regular grid, the dashed one at random amplitudes.  Details in text.  The dotted line shows the difference between the expanded state and a vacuum state truncated to the extent of the grid; the norm of the truncation error is around $10^{-3}$.  The tension is the Frobenius norm of the matrix $T$ defined in the text.
\endinsert

The problem of expanding a state over coherent states is similar to expanding a vector over an overcomplete basis in the plane.  The main difference is that large sets of vectors in the plane are linearly dependent, while sets of coherent states are not.  The two approaches are illustrated below, with an expansion of the vector marked $+$ along the indicated directions.  The black lines show the expansion
$$ \vec x = π^{-1}∑_{r=1}^R (\vec u_r\cdot\vec x)\vec u.$$
As can be seen, this is a very good approximation even with five vectors; in fact, it is hard to visually distinguish the approximate point from $\vec x$ even with 3 directions.  The red lines show $\vec x$ expanded over the two directions closest to it.  The interesting thing is the size of the change of the expansion coefficients when $\vec x$ moves perpendicular to $\vec x$.

\topinsert\XeTeXpicfile circle-1.pdf 
\endinsert

\beginsection{Regularisation of the discrete expansion problem}
%% Picard plots

@@

To capture the concept of a smooth expansion over an irregular set of nonorthogonal vectors, we'll define a quantity called the sampling tension.  Suppose that $|u〉$ and $|v〉$ are nearly parallel vectors, and the state $|ψ〉$ is expanded over a set that includes them, as $|ψ〉≈c_u|u〉+c_v|v〉+⋯$.  A relaxed expansion is one where $c_u|u〉$ and $c_v|v〉$ are as similar as possible; this can be ensured, regardless whether the $c$ are complex numbers of whether $|u〉$ and $|v〉$ are parallel or antiparallel, by requiring that the component of $c_u|u〉$ along $|v〉$ is equal to the component of $c_v|v〉$ along $|u〉$.  The difference between these components is
$$T_{uv}=\left| {c_v〈u|v〉\over ‖|u〉‖} - {c_u〈v|u〉\over ‖|v〉‖}\right|.$$
These differences form a tension matrix.  In the case that $|u〉$ and $|v〉$ are nearly orthogonal, their components should be independent.  In this case, the $T_{uv}$ will be small, because of the proportionality to $〈u|v〉$.  The tension matrix is a reasonable measure of the smoothness of an expansion: its Frobenius norm, the sum in quadrature of the components, would be a good summary.

The idea is that a relaxed expansion has all the coefficients pulling in the same direction, but an expansion with large tension is a tug-of-war, with adjacent coefficients pulling the expanded state in opposite ways.

The tensions were calculated for the grid and irregular expansions above, and plotted in the second graph.  Despite the much worse convergence of random sampling, the weights have very nearly the same tension as the regular grid.

@@
